---
title: "Landslide Density"
author: "Dan"
date: today
format: 
  html:
    code-fold: true
  docx: default
  pdf: default
editor: visual
execute:
  warning: false
bibliography: references.bib
---

# Outline

- Definition
  - Number, area, volume per unit area by landform type
  - Number, area, volume per unit area as functions of terrain elements
- RILs, virtual watershed
  - Density by RIL type
- Geomorphons?
  - Density by landform type
- Terrain Elements
  - Partial contributing area
  - Gradient
  - Factor of Safety
  - Curvature
  - Geology
  - Soils
  - Land cover
  - Precipitation
  - Landslide density as a function of terrain elements
    - ROC, AUC
    - Multivariate, GAM, ...?
- 

# Introduction {#sec-Intro}

In previous sections, we have provided methods to align two DTMs, to determe an appropriate elevation-difference level-of-detection, and to use lidar-change-detection to delineate a set of landslide-scar polygons. Now we want to develop methods using these polygons to map landslide density across the landscape. We will look at variations in landslide density across landform types, including certain RILs, and as a function of terrain elements, including topographic attributes, substrate, land cover, and precipitation. Density provides a measure of landslide susceptibility, in this case, the likelihood of encountering a landslide scar, since we are basing measures of density on past landslide occurrences identified with lidar change detection. We will also translate measures of density to determine where across the landscape most landslides occurred.

## Density

We define "landslide density" in three ways:

1)  the number of landslides per unit area,
2)  the area of landslide scars per unit area,
3)  the volume of landslide scars per unit area.

We can compare these different measures of density across landform types, e.g., what is the density of landslides in bedrock hollows versus planar hillslopes? We can also calculate density as a function of terrain elements, e.g., how does landslide density vary with slope gradient? Landforms are also defined as functions of terrain-element values, so this combination of perspectives will help in evaluating potential modifications of landform and RIL definitions that might better differentiate more and less landslide-prone terrain.

## Landslides

```{r}
#| label: Set path
#|  
path <- "c:\\work\\data\\postmortem\\"
dem_name <- "elev"
dem_id <- "smooth"
data_id <- "06_17"
executable_path <- "c:\\work\\sandbox\\"
scratch_dir <- "c:\\work\\scratch"
NOFILE <- "NOFILE"

```

```{r}
#| label: LShunter
# Identify files to use, set parameter values

library(TWutils)

dem <- paste0(path, dem_name, "_", dem_id)
accum <- paste0(path, "accum_", dem_id)
outlier <- paste0(path, "outlier_", data_id)
roads <- paste0(path, "roads")
in_grad <- paste0(path, "grad_15")
threshold1 <- -5.0
threshold2 <- -2.5
min1 <- 0.4
min2 <- 0.3
max_accum1 <- 1000.
max_accum2 <- 1000.
minSize <- 10. # minimum patch size, see figures above
grad_length <- 15.
road_buffer <- 20.
out_grad <- "NOFILE"
out_patch <- paste0(path, "patch_", data_id)
executable_dir <- paste0(executable_path, "landslideutilities\\projects\\LShunter\\x64\\release")

returnCode <- TWutils::LShunter(
  dem,
  outlier,
  threshold1,
  threshold2,
  in_grad,
  min1,
  min2,
  accum,
  max_accum1,
  max_accum2,
  roads,
  road_buffer,
  minSize,
  grad_length,
  out_grad,
  out_patch,
  scratch_dir,
  executable_dir) 

if (returnCode != 0) {
  return(returnCode)
  stop("Error in LShunter")
}

```

## Landforms

```{r}
#| label: RIL
# Model landforms

library(TWutils)

out_RIL <- paste0(path, "RIL_", data_id)
# Check for existing
if (!file.exists(paste0(out_RIL, ".flt"))) {
  # set parameter values
  dem <- paste0(path, dem_name, "_", dem_id)
  radius <- 7.5
  closest_node <- c(`NUM WIDTHS` = 0,
                   `MAX RADIUS` = 250,
                   `MIN RADIUS` = 250)
  as2_threshold <- 1000.
  plan_curvature_threshold <- 0.005
  gradient_threshold <- 0.5
  fluvial_area_threshold <- 0.07
  valley_depth_max <- 5.0
  valley_buffer <- 2.5
  valley_max_hole <- 1000.
  valley_min_patch <- 20.
  hollow_gradient <- c(PRIMARY = 0.3,
                       SECONDARY = 0.25,
                       DIF1 = 1.0,
                       DIF2 = 0.45,
                      `MIN POLY GRAD` = 0.2)
  hollow_tangential <- c(PRIMARY = -0.005,
                         SECONDARY = -0.02,
                        `MIN POLY TAN` = 0.0)
  hollow_profile <- c(PRIMARY = -0.01,
                      SECONDARY = -0.06,
                     `MIN POLY PROF` = -0.015)
  grad_proportion <- c(GRAD = 0.5, PROP = 0.1)
  tan_proportion <- c(TAN = 0.0, PROP = 0.5)
  fill_hollow_embayments <- TRUE
  hollow_area_threshold <- 5.
  hollow_max_hole <- 1000.
  hollow_min_patch <- c(INITIAL = 0, FINAL = 300)
  gorge_gradient <- c(PRIMARY = 0.7,
                     SECONDARY = 0.45,
                     DIF1 = 0.95,
                     DIF2 = 0.7)
  gorge_max_hole <- 1000.
  gorge_min_patch <- 0
  slope_thresholds <- c(SLOPE = 0.4,
                        STEEP = 0.7)
  curve_thresholds <- c(CONVERGENT = 0.01,
                        DIVERGENT = -0.01)
  edge_smoothing_iterations <- 11
  road_shapefile <- NOFILE
  road_buffer <- 15.
  in_closest_node <- NOFILE
  in_dist_to_channel <- NOFILE
  in_drainage_wing <- NOFILE
  in_valley_floor <- NOFILE
  in_upgrad <- NOFILE
  in_downgrad <- NOFILE
  in_tangential <- NOFILE
  in_plan <- NOFILE
  out_closest_node <- paste0(path, "closestNode_", data_id)
  out_dist_to_channel <- paste0(path, "distToChan_", data_id)
  out_drainage_wing <- paste0(path, "wings_", data_id)
  out_valley_floor <- paste0(path, "valley_", data_id)
  out_upgrad <- paste0(path, "upgrad_", data_id)
  out_downgrad <- paste0(path, "downgrad_", data_id)
  out_tangential <- paste0(path, "tan_", data_id)
  out_plan <- paste0(path, "plan_", data_id)
  out_nodes <- NOFILE
  out_zero_order <- NOFILE
  attribute_list <- ril_default_attributes("c:\\work\\data\\postmortem\\prism_wasp_m")
  overwrite <- TRUE
  executable_dir <- paste0(executable_path, "channelutilities\\projects\\RIL\\x64\\release")

  ril_return <- TWutils::RIL(dem,
                      scratch_dir,
                      out_RIL,
                      radius,
                      closest_node,
                      as2_threshold,
                      plan_curvature_threshold,
                      gradient_threshold,
                      fluvial_area_threshold,
                      valley_depth_max,
                      valley_buffer,
                      valley_max_hole,
                      valley_min_patch,
                      hollow_gradient,
                      hollow_tangential,
                      hollow_profile,
                      grad_proportion,
                      tan_proportion,
                      fill_hollow_embayments,
                      hollow_area_threshold,
                      hollow_max_hole,
                      hollow_min_patch,
                      gorge_gradient,
                      gorge_max_hole,
                      gorge_min_patch,
                      slope_thresholds,
                      curve_thresholds,
                      edge_smoothing_iterations,
                      road_shapefile,
                      road_buffer,
                      in_closest_node,
                      in_dist_to_channel,
                      in_drainage_wing,
                      in_valley_floor,
                      in_upgrad,
                      in_downgrad,
                      in_tangential,
                      in_plan,
                      out_closest_node,
                      out_dist_to_channel,
                      out_drainage_wing,
                      out_valley_floor,
                      out_upgrad,
                      out_downgrad,
                      out_tangential,
                      out_plan,
                      out_nodes,
                      out_zero_order,
                      attribute_list,
                      overwrite,
                      executable_dir=executable_dir)
}
```
