---
title: "Mosaic_Reproject_DTM"
author: "Elise Freeman"
format: html
editor: visual
---

## Mosaic & Reproject DTM

Mosaic DTM tiles (downloaded from lidar portal), convert vertical units to meters, and reproject to NAD 1983 HARN StatePlane Washington South FIPS 4602 (Meters) \[EPSG:2856\]

Requires terra package \[install.packages("terra")\]

## Inputs

```{r}
library(terra)
# Folder containing the individual DTM tiles
input_dir   <- "path/to/DTMs"
# File pattern to match your tiles
file_pattern <- "\\.tif$"

# Output folder
output_dir <- "path/to/output/location"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Target CRS: NAD83(HARN) / Washington South (Meters)
target_crs <- "EPSG:2856"

# Vertical unit conversion factor.
# If DTM is already in meters, set to 1 (no-op) and skip the calc step.
ft_to_m <- 0.3048   
```

## Mosaic tiles & convert vertical units

Can specify mosaic operator in bottom code chunk if you want something other than mean/blend.

```{r}
tile_files <- list.files(input_dir, pattern = file_pattern,
                         full.names = TRUE, ignore.case = TRUE)

if (length(tile_files) == 0) {
  stop("No tiles found in input_dir matching file_pattern. Check the path/pattern.")
}


# Read all tiles into a SpatRasterCollection (better to use for large number of tiles)
tile_list <- lapply(tile_files, rast)
rsrc <- sprc(tile_list)

# Mosaic. fun options: "mean" (default, blends overlaps), "min", "max",
# "median", "first" (use first raster's value where tiles overlap)
message("Mosaicking tiles...")
dtm_mosaic <- mosaic(rsrc, fun = "mean")

names(dtm_mosaic) <- "elevation"
```

Then convert vertical units

```{r}
dtm_mosaic_m <- dtm_mosaic * ft_to_m
```

Reproject to NAD83 Harn Stateplane WA South FIPS 4602 (meters) with bilinear resampling

```{r}
dtm_final <- project(dtm_mosaic_m, target_crs, method = "bilinear")
```

Write output

```{r}
out_path <- file.path(output_dir, "dtm_mosaic_m_epsg2856.tif")

writeRaster(
  dtm_final,
  out_path,
  overwrite  = TRUE,
  datatype   = "FLT4S",              # 32-bit float
  gdal       = c("COMPRESS=LZW", "TILED=YES")
)
```

Quick check to see if it worked

```{r}
print(dtm_final)
plot(dtm_final, main = "Mosaicked DTM (meters) - EPSG:2856")
```
