---
title: "Quantiles"
author: "Dan"
date: today
format: gfm
editor: visual
execute:
  echo: false
bibliography: references.bib
---

Even after alignment of two sequential DTMs, there will be persistent errors in the measured elevation differences. We need to differentiate actual ground-surface-elevation changes from random errors (noise) in the two DTMs. The minimum value above which these differences can reliably be interpreted as changes in elevation is referred to as the level of detection (LoD). Differences less than the LoD are considered random noise. Differences greater than the LoD are recognized as changes in elevation occurring between acquisition of the two lidar datasets. The magnitude of the LoD translates to the level of confidence that can be placed in the eroded and deposited volumes inferred from these measured changes in elevation.

Error statistics compiled for each DTM can be used to estimate a theoretical LoD based on concepts of error propagation [@brasington2003; @lane2003]. These statistics are based on comparison of ground control points with lidar ground-return elevations. Using the standard deviation of elevation errors estimated for each DTM ($\sigma_{2006}$ for the 2006 lidar and $\sigma_{2017}$ for the 2007 lidar), the standard deviation of differences in elevations between the DTMs ($\sigma_{\Delta z}$) is estimated as

$$
\sigma_{\Delta z} = \sqrt{\sigma_{2006}^2 + \sigma_{2017}^2}
$$ {#eq-sigmaDeltaZ}

For the 2006 lidar, $\sigma_{2006}$ is reported as 0.06m. For the 2017 lidar, $\sigma_{2017}$ is reported as 0.04 m in unvegetated areas and 0.09m in vegetated areas. Using the vegetated value, this gives an estimated $\sigma_{\Delta z}$ of about 0.10 m. Assuming that errors follow a normal distribution centered about zero, we might then assume with 95% confidence (two $\sigma_{\Delta Z}$) that differences greater than 0.2m indicate actual changes in elevation.

As shown in the Coregistration document, however, the degree of noise in the DoD varies over length scales of tens to hundreds of meters, with greater variability over forested areas. Confidence as to what level of measured elevation change represents actual ground-elevationd differences between the two DTMs will vary with location across the DoD. We need a method of characterizing DTM differences that accommodates spatial variability in the frequency distribution of difference values and for using that frequency distribution to determine an appropriate LoD.

For coregistration of the two DTMs, we examined the frequency distribution of $\Delta z$ values over the entire difference raster. After filtered out apparent outliers using Tukey's fences, we used the distribution to identify an LoD for delineating stable terrain where elevation changes had likely not occurred. The locational shifts of the 2017 DTM were then calculated to bring the two DTMs into alignment over that stable terrain. We found that the magnitude of the LoD varied with slope gradient.

```{r}
library(data.table)
library(ggplot2)
library(latex2exp)
library(RColorBrewer)


inputPath <- "c:/work/data/postmortem/"
outbins <- paste0(inputPath, "outbins2")
datafile <- paste0(outbins,"_final.csv")
efinal <- as.data.table(read.csv(datafile))
efinal <- efinal[, "Aspect" := as.factor(Azimuth)]
efinal[, "qmin" := Q1 - 1.5*(Q3-Q1)]
efinal[, "qmax" := Q3 + 1.5*(Q3-Q1)]

p_LoD <- ggplot(efinal) + 
  geom_point(aes(x=tan(asin(Slope)), y=qmin, color=Aspect), size=3, alpha = 0.8) +
  geom_smooth(aes(x=tan(asin(Slope)), y=qmin), linewidth = 1, color="black", se=FALSE) +
  geom_point(aes(x=tan(asin(Slope)), y=qmax, color=Aspect), size=3, alpha=0.8) +
  geom_smooth(aes(x=tan(asin(Slope)), y=qmax), linewidth = 1, color="black", se = FALSE) +
  labs(title = TeX(r"(LoD estimated from global frequency distribution of $\Delta z$)"),
       subtitle = '2006 to 2017 lidar DTMs, Post-Mortem study area',
       x = 'Gradient', 
       y='LoD (m)') +
  theme(legend.position="right") +
  scale_color_brewer(palette = 'Set1', 
                     labels=c("N","NE","E","SE","S","SW","W","NW")) +
  scale_x_continuous(labels=scales::percent)
p_LoD
```

Given, however, the increase in uncertainty in elevation differences under forest canopy, this gradient dependence may simply reflect a trend for agricultural fields and pastures to occur in flatter areas and forests on hillslopes. In any case, this LoD based on the observed distribution of $\Delta z$ values is larger than that based on @eq-sigmaDeltaz. We need a method to measure local variability in $\Delta z$ with which to estimate our level of certainty that a difference in elevation between two DTMs represents an actual change in elevation. We can expand on the use of quartiles in measured $\Delta z$ values that we developed to identify stable terrain for coregistration.

Program quantiles builds a frequency distribution of elevation differences over a circular moving window of specified radius. Potential outliers are filtered out, based on Tukey's fences (k=1.5), and the first, second, and third quartiles are calculated. The mean and standard deviation of the elevation differences are also calculated. The z-score is calculated based on the mean and standard deviation. For pixels with values less than the 1^st^ quartile $q_1$, the value $(\Delta z - q_1)/(q_3-q_1)$ is calculated. For pixels with values greater than the 3rd quartile $q_3$, the quantity $(\Delta z - q_1)/(q_3-q_1)$ is calculated. These quantities provide a measure of how extreme the pixel $\Delta z$ value is compared to the entire frequency distribution (minus outliers). The output from program quantiles is a set of rasters for each of these statistics.

```{r}
library(TerrainWorksUtils)
library(stringr)

inputPath <- "c:/work/data/postmortem/"
in_raster <- paste0(inputPath, "difN.flt")
radius <- 25. # radius of the moving circular window
buffer <- 2 # spacing in pixels between window centers
out_outlier <- paste0(inputPath, "outlier_50")
out_q1 <- paste0(inputPath, "q1_50")
out_q2 <- paste0(inputPath, "q2_50")
out_q3 <- paste0(inputPath, "q3_50")
out_mean <- paste0(inputPath, "mean_50")
out_zscore <- paste0(inputPath, "zscore_50")
out_prob <- paste0(inputPath, "prob_50")
scratch_dir <- "c:/work/scratch"
executable_dir <- "c:/work/sandbox/gridutilities/projects/quantiles/x64/release"
program_name <- "quantiles"

returnCode <- TerrainWorksUtils::quantiles(in_raster,
                  radius,
                  buffer,
                  out_outlier,
                  out_q1,
                  out_q2,
                  out_q3,
                  out_mean,
                  out_zscore,
                  out_prob,
                  scratch_dir,
                  executable_dir,
                  program_name)

if (returnCode != 0) {
  stop("Error in quantiles")
}
```

With the output from program quantiles, the expectation is that extreme values are associated with actual changes in elevation between 2006 and 2017. This brings up two questions:

1.  What length scale to use; that is, what radius for the circular window?

2.  How extreme does a difference between the two DTMs need to be to indicate an actual change in elevation?

We'll examine the scale issue first. We want to resolve landslide features, both loss of material in initiation zones and deposition of material downslope. The moving window must be large compared to the areal extent of landslide scars and deposits, otherwise the elevation changes associated with these features will comprise too large a portion of the window area - of the frequency distribution - to be extreme compared to the rest of the window. Conversely, a large window may also overlap zones of high and low variability, thus limiting ability to resolve smaller changes in elevation in low-variability zones.
