---
title: "SusNRunPointCloud_Workflow"
author: "Elise Freeman"
format: html
editor: visual
---

## Point Cloud Pre-Processing Workflow

1.  Filter ground returns (Classification == 2) and write to a GeoPackage as a point feature class

2.  Merge LAZ tiles filtered for ground returns

3.  Compute a point density raster from ground return points (target pixel size = 3 m)

4.  Reproject density (and raw count) rasters to EPSG:2856

Requires: lasR, terra, sf

For terra and sf: install.packages(c("terra", "sf"))

lasR is not on CRAN (by design, not due to archival)

Install with: install.packages("lasR", repos = "[https://r-lidar.r-universe.dev")](https://r-lidar.r-universe.dev%22))

## Inputs

```{r}
library(lasR)
library(terra)
library(sf)

# Folder containing the laZ files
input_dir  <- "path/to/laz_tiles"

# Output folder
output_dir <- "path/to/output"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Merged LAZ path
ground_merged_path <- file.path(output_dir, "ground_merged.laz")

# Path to geopackage output location
gpkg_path  <- file.path(output_dir, "ground_points.gpkg")
gpkg_layer <- "ground_points"

density_tif_native_path <- file.path(output_dir, "ground_density_native_crs.tif")
density_tif_2856_path   <- file.path(output_dir, "ground_density_epsg2856.tif")
count_tif_native_path   <- file.path(output_dir, "ground_count_native_crs.tif")
count_tif_2856_path     <- file.path(output_dir, "ground_count_epsg2856.tif")


target_pixel_size_m <- 3 #in meters
target_crs <- "EPSG:2856"  # NAD83(HARN) / Washington South (Meters)
```

## Filter and merge LAZ

Filters ground returns by tile, the merges into one file

```{r}
las_files <- list.files(input_dir, pattern = "\\.(laz|las)$",
                        full.names = TRUE, ignore.case = TRUE)

if (length(las_files) == 0) {
  stop("No LAS/LAZ files found in input_dir. Check the path.")
}

read_stage   <- reader_las()
filter_stage <- keep_ground()
write_stage  <- write_las(ground_merged_path)   # no "*" -> single merged file

filter_merge_pipeline <- read_stage + filter_stage + write_stage
exec(filter_merge_pipeline, on = las_files)
message("Ground-only merged point cloud written to: ", merged_las_path) #just tells you that it worked and where to find it
```

This detects the native CRS and units of the las file, which are needed for the GeoPackage and density raster.

```{r}
read_probe <- reader_las()
probe_pipeline <- read_probe + lasR::rasterize(1, "count")  # res=1 just to read CRS
probe_rast <- exec(probe_pipeline, on = ground_merged_path)

native_crs <- st_crs(probe_rast)
native_units <- native_crs$units_gdal
if (is.null(native_units)) {
  stop("Could not determine linear units of the LAS file's CRS. ",
       "Check that the point cloud has a valid projected CRS assigned.")
}
message("Detected native CRS units: ", native_units)

if (is_us_survey_feet) {
  native_res <- target_pixel_size_m * (3937 / 1200)   # US survey foot
} else if (is_feet) {
  native_res <- target_pixel_size_m / 0.3048            # international foot
} else {
  native_res <- target_pixel_size_m                     # already metres
}
message("Native-CRS resolution corresponding to ", target_pixel_size_m,
        " m pixels: ", round(native_res, 6), " (", native_units, ")")
```

## Export ground points to GeoPackage

(Classification == 2)

```{r}
extract_points <- function(data) return(data)
read_all <- reader_las()
cb       <- callback(extract_points, expose = "xyzc", no_las_update = TRUE)
read_pipeline <- read_all + cb

ground_df <- exec(read_pipeline, on = ground_merged_path)

if (is.null(ground_df)) {
  stop("exec() returned NULL when reading the ground-only point cloud. ",
       "Check that ground_merged_path exists and is a valid, non-empty ",
       "LAS/LAZ file.")
}
if (is.list(ground_df) && !is.data.frame(ground_df)) {
  message("exec() returned a list; combining chunks into one data.frame...")
  ground_df <- do.call(rbind, ground_df)
}

message("Read ", nrow(ground_df), " ground points. Columns: ",
        paste(names(ground_df), collapse = ", "))

# Sanity check: keep_ground() should mean everything here is class 2, but
# confirm rather than assume.
if ("Classification" %in% names(ground_df)) {
  class_tab <- table(ground_df$Classification)
  message("Classification breakdown of merged ground file:")
  print(class_tab)
  if (!all(names(class_tab) == "2")) {
    warning("Merged ground file contains classes other than 2 -- check ",
            "that keep_ground() behaved as expected.")
  }
}

if (nrow(ground_df) == 0) {
  stop("No ground points found after filtering. Is the point cloud ",
       "already ground-classified?")
}

# Exports ground points to GeoPackage
ground_sf_native <- st_as_sf(ground_df, coords = c("X", "Y"), crs = native_crs)
ground_sf_2856   <- st_transform(ground_sf_native, target_crs)

st_write(ground_sf_2856, gpkg_path, layer = gpkg_layer,
         driver = "GPKG", delete_layer = TRUE)
message("Ground return points written to: ", gpkg_path,
        " (layer: ", gpkg_layer, ")")
```

## Create Point Density Raster

Ground points only. Rasterizes the ground-only points directly with terra, using a template raster at native_res (native CRS units) so density matches the intended 3m pixel size once reprojected.

```{r}
ground_vect <- vect(ground_sf_native)

# Template raster spanning the ground points' extent, at native_res
template <- rast(ground_vect, resolution = native_res)

bg_value <- if (fill_empty_cells_with_zero) 0 else NA
count_rast <- terra::rasterize(ground_vect, template, field = 1, fun = sum,
                               background = bg_value)
names(count_rast) <- "ground_point_count"

n_na <- global(count_rast, fun = "isNA")[1, 1]
n_total <- ncell(count_rast)

#Normalized density (ground points per square meter)
cell_area  <- target_pixel_size_m^2   # true physical cell area, in m^2 instead of native_res^2, which would be in native (e.g. square feet) units and give density in the wrong units
density_rast <- count_rast / cell_area
names(density_rast) <- "ground_point_density"

writeRaster(density_rast, density_tif_native_path, overwrite = TRUE)
message("Native-CRS ground density raster written to: ", density_tif_native_path)

```

# Project Density Raster to ESPG:2856

Resampling is bilinear, can change to "near" if appropriate.

```{r}
density_rast_2856 <- project(density_rast, target_crs, method = "bilinear",
                             res = target_pixel_size_m)
writeRaster(density_rast_2856, density_tif_2856_path, overwrite = TRUE)
message("Reprojected ground density raster written to: ", density_tif_2856_path)
```

Quick check to see if it worked

```{r}
print(density_rast_2856)
print(ground_sf_2856[1:min(5, nrow(ground_sf_2856)), ])
```
